---
title: "finalrep"
author: "Osman Batuhan Şahin"
date: "13 01 2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Including Libraries

```{r warning=FALSE}
library(dplyr)
library(stringr)
library(corrplot)
library(ggplot2)
library(caret)
library(ROSE)
library(factoextra)
library(NbClust)

options(warn=-1)

```


## Data Description

I found my dataset from UCI. You can visit page with this url: https://archive.ics.uci.edu/ml/datasets/diabetes+130-us+hospitals+for+years+1999-2008
This dataset is about diabetes patients of 130 hospitals in the USA between 1999-2008. Contains 101,766 rows and 50 columns.

```{r}
data = read.csv("diabetic_data.csv", stringsAsFactors = TRUE)

head(data)

```


## Missing values
"?" and "Unknown/Invalid" are written for the missing data in the data set. This plots shows them. I replace them with NA.
```{r}
barplot(table(data$weight))
barplot(table(data$race))
barplot(table(data$age))
barplot(table(data$payer_code))
barplot(table(data$medical_specialty))

data = na_if(data, "?")
data = na_if(data, "Unknown/Invalid")
```
Number of missing values for each column.
```{r}
sapply(data, function(x) sum(is.na(x)))
```
weight, payer_code and and medical_specialty columns has a lot of missing values. So deleted.
```{r}
data = select(data, -c(weight, payer_code, medical_specialty))
dim(data)
```

Diagnosis 2, Diagnosis 3 deleted. I only use the primary diagnosis.
```{r}
data = select(data, -c(diag_2, diag_3))
dim(data)
```
Citoglipton and examide deleted because all rows have the same value.
```{r}
table(data$citoglipton)
table(data$examide)

data = select(data, -c(citoglipton, examide))
dim(data)
```

There are 2 csv in data folder: https://archive.ics.uci.edu/ml/machine-learning-databases/00296/ . Second csv show descriptions of admission_type_id, discharge_disposition_id, admission_source_id. I used this csv file in the next steps.


The admission_type_id feature was taking a value between 1-8. The meanings of these values represent "emergency", "elective", "new born" and "not available" values in hospitals. Due to this situation, i re-filled the values in our data according to 4 categories: "emergency", "elective", "new born" and "other".
```{r}
barplot(table(data$admission_type_id))

data$admission_type_id <- case_when(data$admission_type_id %in% c("1","2","7") ~ "Emergency",
                                    data$admission_type_id %in% c("5","6","8") ~ "Other",  
                                    data$admission_type_id %in% c("3") ~ "Elective", 
                                    data$admission_type_id %in% c("4") ~ "Newborn")
data$admission_type_id <- as.factor(data$admission_type_id)

barplot(table(data$admission_type_id))

```

Due to codes in csv file i changed numeric values as "Discharged to home" and "Other".
```{r}
barplot(table(data$discharge_disposition_id))
data$discharge_disposition_id <- case_when(data$discharge_disposition_id %in% c("1","6","8","13","19") ~ "Discharged to home",
                                    data$discharge_disposition_id %in% c("2","3","4","5","7","9","10","11","12","14","15","17","20","21","22","23","24","27","28","29","30","18","25","26") ~   "Other")

data$discharge_disposition_id <- as.factor(data$discharge_disposition_id)


barplot(table(data$discharge_disposition_id))
```

Due to codes in csv file i change numeric values as "Referral", "Other" and "Emergency".
```{r}
barplot(table(data$admission_source_id))

data$admission_source_id <- case_when(data$admission_source_id %in% c("1","2","3") ~ "Referral",
                                    data$admission_source_id %in% c("4","5","6","8","9","10","11","12","13","14","15","17","18","19","20","21","22","23","24","25","26") ~   "Other",  
                                    TRUE ~ "Emergency")    

data$admission_source_id <- as.factor(data$admission_source_id)

barplot(table(data$admission_source_id))
```

Readmitted feature took 3 different values. No readmission("No"), a readmission in more than 30 days(">30"), a readmission in less than 30 days("<30"). I assumed "> 30" and "NO" values as similar cases and converted it to binary class by encoding it as 0-1. Now, this feature tells us a patient readmitted in 30 days or not. Barplot shows this dataset is imbalanced. Imbalancing is a problem for classification.

```{r}
data$readmitted <- case_when(data$readmitted %in% c(">30","NO") ~ "0",
                              TRUE ~ "1")
data$readmitted <- as.factor(data$readmitted)
barplot(table(data$readmitted))

```
Removed duplicated patients. After patient_nbr and encounter_id columns removed.
```{r}
data <- data[!duplicated(data$patient_nbr),]

data = select(data, -c(encounter_id, patient_nbr))

dim(data)
```

## Missing Data imputation
With calc_mode function, i calculated mode and imputed most frequent value for categorical columns race and gender.
```{r}
calc_mode <- function(x){
  
  # List the distinct / unique values
  distinct_values <- unique(x)
  
  # Count the occurrence of each distinct value
  distinct_tabulate <- tabulate(match(x, distinct_values))
  
  # Return the value with the highest occurrence
  distinct_values[which.max(distinct_tabulate)]
}

imputeddata = data %>% 
  mutate(gender = if_else(is.na(gender), 
                         calc_mode(gender), 
                         gender))

imputeddata = data %>% 
  mutate(race = if_else(is.na(race), 
                         calc_mode(race), 
                         race))
dim(imputeddata)
```

diag_1 has missing values. So deleted rows with missing values.

```{r}
imputeddata = na.omit(imputeddata)
dim(imputeddata)
```


This barplot shows "diag_1" column. Diagnosis has ICD codes. I regroup them to categories in primary_diagnosis column.
```{r}
barplot(table(imputeddata$diag_1))
```
I used the list of icd-9 codes for primary_diagnosis column.
```{r warning=FALSE}
imputeddata$diag_1 <- as.character(imputeddata$diag_1)

imputeddata<- mutate(imputeddata, primary_diagnosis =
                                          ifelse(str_detect(diag_1, "V") | str_detect(diag_1, "E"),"Other", 
                                          # disease codes starting with V or E are in “other” category;
                                          ifelse(str_detect(diag_1, "250"), "Diabetes",
                                          ifelse((as.integer(diag_1) >= 390 & as.integer(diag_1) <= 459) | as.integer(diag_1) == 785, "Circulatory",
                                          ifelse((as.integer(diag_1) >= 460 & as.integer(diag_1) <= 519) | as.integer(diag_1) == 786, "Respiratory", 
                                          ifelse((as.integer(diag_1) >= 520 & as.integer(diag_1) <= 579) | as.integer(diag_1) == 787, "Digestive", 
                                          ifelse((as.integer(diag_1) >= 580 & as.integer(diag_1) <= 629) | as.integer(diag_1) == 788, "Genitourinary",
                                          ifelse((as.integer(diag_1) >= 140 & as.integer(diag_1) <= 239), "Neoplasms",  
                                          ifelse((as.integer(diag_1) >= 710 & as.integer(diag_1) <= 739), "Musculoskeletal",          
                                          ifelse((as.integer(diag_1) >= 800 & as.integer(diag_1) <= 999), "Injury",                    
                                          "Other"))))))))))
table(imputeddata$primary_diagnosis)
```
Removed "diag_1". Make type of primary_diagnosis factor.
```{r}
imputeddata$primary_diagnosis <- as.factor(imputeddata$primary_diagnosis)
imputeddata$diag_1 <- NULL
```

## Splitting Data

Imputed data splitted into training and test sets with 80:20 proportion.

```{r}
set.seed(42)
train <- createDataPartition(imputeddata$readmitted, p = 0.8, list = FALSE)
training <- imputeddata[train, ]
testing <- imputeddata[-train, ]
table(training$readmitted)
```
## Balancing Dataset

I used "ROSE" (Random Over Sampling Examples) method for balancing dataset. ROSE method uses undersampling for class 0 (“>30” and “No”) and uses bootstrapping to increase artificial samples for class 1 (“<30”) to make the dataset balanced.

```{r}
data_rose = ROSE(readmitted ~., data = training)$data
table(data_rose$readmitted)
```
## Logistic Regression

I applied binary logistic regression to balanced and imputed data because "readmitted" column has binary values. Logistic Regression is the appropriate regression analysis to conduct when the dependent variable is binary.

```{r warning=FALSE}
lreg<-train(readmitted~.,data=data_rose, method="glm",family="binomial")
```
Confusion matrix shows model has 84.31% accuracy. 

```{r warning=FALSE}
lreg_pred <- predict(lreg, testing)

confusionMatrix(lreg_pred, testing$readmitted)
```
Variable Importances
```{r}
varImp(lreg)
```

## Decision Tree

I applied decision tree classification to balanced and imputed data. Decision trees are one of the first to come to mind and most reliable algorithms for classification. 

```{r}
dt<-train(readmitted~.,data=data_rose, method="rpart")
```
Confusion matrix shows accurary of model is 87.62%.
```{r}
dt_pred <- predict(dt, testing)

confusionMatrix(dt_pred, testing$readmitted)

```


## Decision Tree With Imbalanced Class

```{r}
imdt<-train(readmitted~.,data=training, method="rpart")
```
```{r}
imdt_pred <- predict(imdt, testing)

confusionMatrix(imdt_pred, testing$readmitted)

```
If I compare statistics of the balanced and imbalanced models accuracy of imbalanced model is better. But accuracy score is not a good metric for imbalanced classification. With imbalanced model, specificity is too low and sensitivity is too high that is not good.

## Kmeans Clustering
```{r}
kmodel <- kmeans(imputeddata[c("time_in_hospital", "num_lab_procedures", "num_procedures", "number_outpatient", "number_emergency", "num_medications", "number_inpatient", "number_diagnoses")], centers = 2)
km_cluster <- data.frame(imputeddata[c("time_in_hospital", "num_lab_procedures", "num_procedures", "number_outpatient", "number_emergency", "num_medications", "number_inpatient", "number_diagnoses")],
  cluster = as.factor(kmodel$cluster)
)

head(km_cluster)

(BSS <- kmodel$betweenss)
(TSS <- kmodel$totss)
BSS / TSS * 100
```
```{r}
kmodel3 <- kmeans(imputeddata[c("time_in_hospital", "num_lab_procedures", "num_procedures", "number_outpatient", "number_emergency", "num_medications", "number_inpatient", "number_diagnoses")], centers = 3)
km_cluster3 <- data.frame(imputeddata[c("time_in_hospital", "num_lab_procedures", "num_procedures", "number_outpatient", "number_emergency", "num_medications", "number_inpatient", "number_diagnoses")],
  cluster = as.factor(kmodel3$cluster)
)

head(km_cluster3)

(BSS <- kmodel3$betweenss)
(TSS <- kmodel3$totss)
BSS / TSS * 100
```
```{r}
kmodel4 <- kmeans(imputeddata[c("time_in_hospital", "num_lab_procedures", "num_procedures", "number_outpatient", "number_emergency", "num_medications", "number_inpatient", "number_diagnoses")], centers = 4)
km_cluster4 <- data.frame(imputeddata[c("time_in_hospital", "num_lab_procedures", "num_procedures", "number_outpatient", "number_emergency", "num_medications", "number_inpatient", "number_diagnoses")],
  cluster = as.factor(kmodel4$cluster)
)

head(km_cluster4)

(BSS <- kmodel4$betweenss)
(TSS <- kmodel4$totss)
BSS / TSS * 100
```
I calculated the quality of the partition for 2 center, 3 center, 4 center respectively.
```{r}
fviz_cluster(kmodel, imputeddata[c("time_in_hospital", "num_lab_procedures", "num_procedures", "number_outpatient", "number_emergency", "num_medications", "number_inpatient", "number_diagnoses")], ellipse.type = "norm")
```
```{r}
fviz_cluster(kmodel3, imputeddata[c("time_in_hospital", "num_lab_procedures", "num_procedures", "number_outpatient", "number_emergency", "num_medications", "number_inpatient", "number_diagnoses")], ellipse.type = "norm")
```
```{r}
fviz_cluster(kmodel4, imputeddata[c("time_in_hospital", "num_lab_procedures", "num_procedures", "number_outpatient", "number_emergency", "num_medications", "number_inpatient", "number_diagnoses")], ellipse.type = "norm")
```
I plot clusters by using the fviz_cluster() function for 2 center, 3 center, 4 center respectively. With this plots and quality scores I can say 2 or 3 clusteri is enough.



